Bayesian latent class model with sparse correlation for etiology estimation

### 2017-01-26 Reports/draft.v1/draft_latentquadexp.tex
report update

\title{Baysian latent class model with partially informative priors for etiology estimation}   % type title between braces
\author{Detian Deng}         % type author(s) between braces
\date{\today}    % type date between braces
\maketitle

@@

\subsection{The Joint Density}
With the specification of likelihood and prior, we can construct the joint density needed for building the MCMC algorithm. Let $G_i$ be an indicator of whether subject $i$ has GS measurements, and define the following three index sets.
\begin{align*}
I_1 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=1 \Big\}\\
I_2 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=0 \Big\}\\
I_3 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=0 \Big\}
\end{align*}
then we can define the joint density of data and parameters as follow by combining all building blocks together:
\begin{align*}
& P(M,\Theta, \eta, \gamma, \delta) \\
= & \prod_{i\in I_1} P(M_i^{GS},M_i^{SS},M_i^{BS} | \Theta, \eta, \gamma, \delta) 
\prod_{i\in I_2} P(M_i^{SS},M_i^{BS} |\Theta, \eta, \gamma, \delta)
\prod_{i\in I_3} P(M_i^{BS} |\gamma, \delta) P(\gamma, \delta)
\end{align*}
\\
From the form of this model, we can see that the model we proposed incorporates data from all the sources with measurements of varying quality. With the joint density fully specified, we can build up a Metropolis-Hastings (within Gibbs) algorithm to simulate from the posterior distribution. The details of the algorithm can be found in the appendix.

\newpage
\section{Simulation Studies}
Three set of simulation studies are carried out to empirically evaluate the effectiveness of our proposed method. For all three sets of studies, we assume there are five candidate pathogens and two binary covariates. And in each simulated data set, there are $500$ case subjects and $1000$ control subjects.

\subsection{Study I}
In the first set of studies, the true lung infection status $L$ of case patients are generated by a Quadratic Exponential Model, where the first order canonical parameters are dependent on both covariates with an interaction effect, and the second order parameters are independent of the covariates.
\begin{align*}
\theta^{(1)}_{ik} = \beta_{k0} + \beta_{k1} x_{i1} + \beta_{k2} x_{i2} + \beta_{k3} x_{i1} x_{i2}
\end{align*}

Also, for $K = 5$, there are $10$ second order parameters, we assume that two of them are zero, which represents that two particular pairs of pathogens infect lungs independently from each other; and the rest eight association parameters share the same negative value: $-1.5$, which stands for the competition among pathogens. Then, the BS and SS measurements for case subjects are simulated based on formula (8) and (9) respectively, and the SS measurements for control subjects are simulated based on formula (10) assuming that there is no infection at all in control patients' lungs. The actual parameter values used in the simulation process are summarized in table \ref{•} in the Appendix. \\

The above data generation procedure is repeated $200$ times, so that we have $200$ independent data sets generated using the same mechanism. Our proposed model is applied to each of these data sets with a sequence of different prior specifications. These specifications share the same values for hyper-parameters $a_k, b_k, c_k, d_k, e_k$ and $f_k$, $k = 1, 2, \ldots, 5$,  because in practice they should be determined by the same interval matching procedure. They also share the same values for $\sigma_1$, and $\sigma_2$, which are set relatively large to represent non-informativeness. And these shared hyper-parameter values are shown in table \ref{tab: }. The difference between these specifications lies in the values of $\alpha_2, \lambda_a$ and $\lambda_b$, which represent the prior belief on the magnitude of pairwise competitions.\\

For each unique prior specification, the posterior means are collected to construct an approximate sampling distribution of our estimator. The mean of these approximate sampling distributions implies empirically the values that our parameter estimates converge to. Thus the overall accuracy of our method is evaluated based on these sampling distributions means. Note that a five-dimensional multivariate binary distribution can be represented by a multinomial distribution with $32$ cells. Let $q_{j}$, $j = 1, \ldots, 32$ be the true multinomial cell probabilities, and let $\hat{q}_j$ be our cell probability estimations based on the sampling distribution means, then the Bhattachayya coefficient \cite{bhattachayya1943measure}, $\sum_{j=1}^{32}\sqrt{q_j \hat{q}_j}$, indicates the general accuracy of our method. \\

In this study, our method is compared against the Bayesian partially-Latent Class Model (pLCM) \cite{wu2015partially}.

Table \ref{tab: } contains the estimation accuracy of our method under different prior specifications, as well as the accuracy of

\subsection{Study II}

\subsection{Study III}
A simulation study is done to explore the effectiveness of our proposed method. Based on the parameter values shown in table \ref{tab: truepar}, a set of 3-dimensional multivariate binary measurement data are simulated. In total, 1000 controls and 1000 cases are simulated.  For each control, only BS measurements are sampled. For cases, they all have SS and BS measurements and $3\%$ of them have additional GS measurements simulated. \\

A sequence of experiments are carried out based on the simulated data set. First, with fixed $3\%$ of GS measurements and $1000$ controls, we sample the posterior distribution with case sample size $200$, $400$, $600$, and $1000$ respectively. Then, following the same sequence of sample sizes, we sample the posterior distribution without using GS measurements. Figure \ref{simu} summarizes the prior and posterior distributions of $(\mu, \pi)$ from these experiments. Prior distributions used for other parameters are plotted in figure \ref{betaprior}, as we can see, aach prior is tuned to be weakly informative, that is, to have relatively large variance and to center at wrong values that are not too far away from the truth.\\

The results of this simulation study suggest that only with a small amount of direct (GS) measurements, the posterior distribution is converging to the true value as the number of observation grows. Also, the posterior credible region has a high coverage probability over the true value even when direct measurements are completely absent. Since the priors for sensitivities and specificities were set to be weakly informative, the robustness of this model is empirically testified.

>>>>

\title{Bayesian latent class model with sparse correlation for etiology estimation}   % type title between braces
\author{Detian Deng}         % type author(s) between braces
\date{\today}    % type date between braces
\maketitle

@@

\subsection{The Joint Density}
With the specification of likelihood and prior, we can construct the joint density needed for building the MCMC algorithm. Let $G_i$ be an indicator of whether subject $i$ has GS measurements, and define the following three index sets.
\begin{align*}
I_1 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=1 \Big\}\\
I_2 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=1 \text{ and } G_i=0 \Big\}\\
I_3 = & \Big \{i \in \{1,2,\ldots, n\}: Y_i=0 \Big\}
\end{align*}
then we can define the joint density of data and parameters as follow by combining all building blocks together:
\begin{align*}
& P(M,\Theta, \eta, \gamma, \delta) \\
= & \prod_{i\in I_1} P(M_i^{GS},M_i^{SS},M_i^{BS} | \Theta, \eta, \gamma, \delta) 
\prod_{i\in I_2} P(M_i^{SS},M_i^{BS} |\Theta, \eta, \gamma, \delta)
\prod_{i\in I_3} P(M_i^{BS} |\gamma, \delta) P(\gamma, \delta)
\end{align*}
\\
From the form of this model, we can see that the model we proposed incorporates data from all the sources with measurements of varying quality, and its likelihood for the latent variable adopts a quadratic exponential model with sparse correlation structure, so we will refer to this model as the Latent Sparse Correlation (LSC) model in the rest of this paper. With the joint density fully specified, the posterior distribution of our LSC model can be sampled by a Metropolis-Hastings (within Gibbs) algorithm, of which the details are described in the appendix.

\newpage
\section{Simulation Studies}
\subsection{Data Simulation}
Three sets of simulation studies are carried out to empirically evaluate the effectiveness of the LSC model under different situations. For all three sets of studies, we assume there are five candidate pathogens and two binary covariates. In each study, $200$ independent data sets are simulated, and in each simulated data set, there are $500$ case subjects and $1000$ control subjects.\\

At the data simulation stage, we first simulate the true lung infection status, then generate the BS and SS measurements. In study I, multiple pathogens are allowed to infect the lung at the same time, and the measurements are of relatively low quality, that is, lower true positive rates and higher false positive rates. In study II, infection is assumed to be caused by a single pathogen, and the measurement quality is the same as in study I. In study III, the true lung status is generated in the same way as in study I, but the measurements have relatively high quality. Details of the study design are described below.

\begin{itemize}
\item[{\bf I}]
In the first set of studies, the true lung infection status $L$ of case patients are generated by a Quadratic Exponential Model, where the first order canonical parameters are dependent on both covariates with an interaction effect, and the second order parameters are independent of the covariates.
\begin{align*}
\theta^{(1)}_{ik} = \beta_{k0} + \beta_{k1} x_{i1} + \beta_{k2} x_{i2} + \beta_{k3} x_{i1} x_{i2}
\end{align*}

Also, for $K = 5$, there are $10$ second order parameters, we assume that two of them are zero, which represents that two particular pairs of pathogens infect lungs independently from each other; and the rest eight association parameters share the same negative value: $-1.5$, which stands for the competition among pathogens. Then, the BS and SS measurements for case subjects are simulated based on formula (8) and (9) respectively, and the SS measurements for control subjects are simulated based on formula (10) assuming that there is no infection at all in control patients' lungs, where $\text{TPR}^{(SS)} \approx 0.1,  \text{TPR}^{(BS)} \approx 0.6, \text{FPR}^{(BS)} \approx 0.45$. And the actual parameter values used in the simulation process are summarized in table \ref{•} in the Appendix.

\item[{\bf II}]
In the second set of studies, the true lung infection status $L$ of case patients are generated by a Multinomial Model, which is equivalent to the above QE model with all the second order parameters set to negative infinity. And the BS and SS measurements are generated in the same way as they are in study I.

\item[{\bf III}]
In the second set of studies, the true lung infection status of case patients are simulated from the same model as in study I, but the parameters that control the measurement quality are set differently, that is $\text{TPR}^{(SS)} \approx 0.8,  \text{TPR}^{(BS)} \approx 0.9, \text{FPR}^{(BS)} \approx 0.1$. Their actually values are summarized in table \ref{•}.\\
\end{itemize}

\subsection{Model Specifications}
In each of the above situation, $200$ independent data sets are generated. The LSC model is applied to each data set (without using GS measurements) with a series of different prior specifications on $\alpha_2, \lambda_a$ and $\lambda_b$, which represent the experts' prior knowledge on the magnitude of the competitions between pathogens. Within each study, the prior specifications share the same values for hyper-parameters $a_k, b_k, c_k, d_k, e_k$ and $f_k$, $k = 1, 2, \ldots, 5$, which control the prior input on measurement quality. Wu, et al.(2015) \cite{wu2015partially} had discussed the model sensitivity to these hyper-parameters and the partial identifiability issue, which also applies to our method, thus we do not further study the sensitivity issue on these hyper-parameters. Their values are selected according to experts' knowledge on the quality of BS and SS measurements, and the true TPR and FPR values are set to be covered by the prior 95\% credible interval. For all three studies, same values are used for $\sigma_1$, and $\sigma_2$, which are large enough to represent non-informativeness. These hyper-parameter values are listed in table \ref{tab: }. \\

For each unique prior specification in each study, we have $200$ sets of posterior samples produced by the LSC model. Their posterior means are collected to construct an approximate sampling distribution of the estimator. The mean of these approximate sampling distributions implies empirically the values that our parameter estimates converge to. Thus the overall accuracy of the LSC model is evaluated based on these sampling distributions means. Note that a five-dimensional multivariate binary distribution can be represented by a multinomial distribution with $32$ cells. Let $q_{j}$, $j = 1, \ldots, 32$ be the true multinomial cell probabilities, and let $\hat{q}_j$ be the cell probability estimations based on the sampling distribution means, then the Bhattachayya coefficient \cite{bhattachayya1943measure}, $\sum_{j=1}^{32}\sqrt{q_j \hat{q}_j} \in [0, 1]$, which measures the similarity between two discrete distributions, is a good metric of the general accuracy of the LSC model. \\

In study I and II, the LSC model is compared against the Bayesian partially-Latent Class Model (pLCM) \cite{wu2015partially}. The pLCM, which originally only considers single-pathogen infection, now allows multi-pathogen infection in its latest release by making it possible to manually specify candidate pathogen combinations and a Dirichlet prior with equal weights. But the saturated model, in which all possible combinations are included, is unstable and lack of identifiability with only a few hundred case subjects. Thus, rather than the saturated model, two most commonly used pLCM specifications are applied: 1) the classic pLCM (pLCM-1) where only single pathogen infections are allowed and 2) the new pLCM (pLCM-2) where not only single pathogen but also all pairs of pathogen infections are allowed.\\

\subsection{Results}
Table \ref{tab: } lists the Bhattachayya coefficients of the LSC model under different prior specifications, and of the two pLCMs in Study I. The same information is visualized in figure \ref{}. As we can see, when the true data generating mechanism allows multi-pathogen infection, the classic pLCM (pLCM-1) performs the worst, the pLCM-2 is the second worst, and the LSC model, across all prior specifications, shows a significant amount of improvement over the two pLCM models. The variations caused by different prior specifications is relatively small comparing to the improvement over pLCM models, especially the single-pathogen model.\\

Model performances in Study II is summarized in table \ref{tab: } and plotted in figure \ref{}. The figure shows that when the true data generating mechanism is actually single pathogen infection only, the pLCM-1 has the best estimation accuracy since that is the true model, and the pLCM-2 performs the worst because it tends to attribute the cause of disease to two-pathogen infection. The LSC model, on the other hand, shows an increasing trend in estimation accuracy as the prior mean of $\theta_2$ gets more negative, because $\theta_2 = -\infty$ makes the model equivalent to the true model. And in practice, as we can see in figure \ref{}, if the true model is not known, the LSC model can still provide an estimate almost as accurate as the pLCM-1 with a moderately negative (e.g. between $-10$ and $-5$) prior mean of $\theta_2$.\\

It is very intuitive to think that better measurement quality leads to more accurate etiology estimation. Figure \ref{} , which compares the etiology estimations, $\hat{\mathbb{E}}(L|Y=1,X)$, from study III and two scenarios of study I, confirms this hypothesis. In this figure, the sampling distributions of the etiology estimations for pathogen A, B and C in the upper panel have larger standard deviations than those in the mid panel, their shapes are more skewed and the estimation bias tends to be larger, and such distributions for pathogen D and E are almost the same in the upper and mid panel. Comparing the lower panel to the mid panel, it is clear that the sampling distributions estimated from high quality measurements have much smaller standard deviations and basically no estimation bias.\\

\begin{center}
\includegraphics[scale=0.45]{DataDescription.png}
\captionof{figure}{{\bf Data Description}: }
\label{data}
\end{center}

\section{PERCH Data Analysis}
As what has been introduced in the first section of this paper, the PERCH study enrolled about 4200 children hospitalized for severe/very severe pneumonia and approximately 5300 controls randomly select from communities across $7$ sites around the world. To demonstrate the application of the LSC model for the analysis of PERCH study data, only the Kenya site data, where there is good availability of both BS and SS measurement data, is used so that the site-specific effect if not a concern.

\section{Conclusion}
